---
title: "Day 2 lecture"
output: html_document
---

```{r}
library(here)
library(readr)
generation <- read.csv(here::here("data/ca_energy_generation.csv"), stringsAsFactors = F )
imports <- read.csv(here::here("data/ca_energy_imports.csv"), stringsAsFactors = F)
```

```{r}
str(generation)
```

```{r}
class(generation$datetime)
class(imports$datetime)
```

## Lubridate package

- The best way to deal with date-time data is to use the lubridate package
- You can convert character variables into datetime format using the as_datetime function
- One advantage of readr::read_csv is that it will often detect and convert datetime variables when importing

```{r}
library(lubridate)
generation$datetime <- as_datetime(generation$datetime)
class(generation$datetime)
```

```{r}
head(generation$datetime)
```

```{r}
imports$datetime <- as_datetime(imports$datetime)
class(imports$datetime)
head(imports$datetime)
```

## reshape2
1. melt –> make data long
2. dcast –> make data wide
3. recast–> melt then cast data
```{r}
head(generation)
```

Right now, the generation dataframe has several observations per row
```{r}
library(reshape2)
long_gen <- melt(generation, id.vars = "datetime",
                             variable.name = "source",
                             value.name = "usage")
head(long_gen)
```


```{r}
head(long_gen[order(long_gen$datetime), ])
```

## Merging
1. merge(x, y, by = c("id", "year"))

x: first dataframe
y: second dataframe
by: variables to match (must have common name)


2. merge(x, y, by.x = "id", by.y = "cd", all.x = T, all.y = T)

Use by.x and by.y if the dataframes have different variable names
Use all.x = T if you want to keep all the observation in the first dataframe (unmatched observations in y are dropped!)
Use all.y = T if you want to keep all observations in the second dataframe (umatched observations in x are dropped!)
Use both (or, simply all = T) to keep all observations!
By default R will drop unmatched observations from both data frames
```{r}
merged_energy <- merge(generation, imports, by = "datetime")
dim(merged_energy)
```
```{r}
head(merged_energy)
```

Our merged dataframe is still wide and untidy
Create a long version called long_merged_energy
```{r}
long_merged_energy <- melt(merged_energy, id.vars = "datetime",
                           variable.name = "source",
                           value.name = "usage")
head(long_merged_energy)
```


## Two approaches to data manipulation
dplyr, and data.table, are two packages used to process data in tabular form

dplyr: chain together a few, simple functions
data.table: compact, syntax-based commands

data.table is much faster on large data (dplyr is explicitly meant for small-to-medium data)
```{r}
library(tidyverse)
```

### dplyr verbs
select –> subset variables
filter –> subset observations based on conditions
mutate –> add new variables
summarize –> reduce multiple observations to a single value (e.g., find the mean)

**Select**
1. select by name:
select(gapminder, continent, pop)
2. select by position:
select(df, c(1, 3, 10))
3. select by range:
select(df, country:pop) or select(df, 1:3)
4. drop variables with -
select(df, -gdpPercap)
```{r}
tmp <- select(merged_energy, biogas, biomass, geothermal, solar)
names(tmp)

tmp <- select(merged_energy, -biogas, -biomass, -geothermal, -solar)
names(tmp)
```

There are several “select helpers” that make subsetting variables very easy
one_of(), contains(), starts_with(), ends_with(), matches()

```{r}
tmp <- select(merged_energy, contains("hydro"), starts_with("bio"))
names(tmp)
```

**Filter**
```{r}
tmp <- filter(merged_energy, imports > 7000)
nrow(tmp)
head(tmp)
```

Multiple conditions
```{r}
tmp <- filter(merged_energy, imports > 7000, natural_gas < 7000)
nrow(tmp)
head(tmp)
```


**Mutate**
mutate creates new variables
```{r}
tmp <- mutate(long_merged_energy, log_usage = log(usage))
head(tmp)
```

**Summarize**
summarize reduces observations to a single value based on functions - mean, sum, sd, min, max, etc.
```{r}
# total energy consumption
summarize(long_merged_energy, total = sum(usage, na.rm = T))
# mean energy consumption
summarize(long_merged_energy, mean_cons = mean(usage, na.rm = T))
```

### A new operator: the pipe %>%
The powerful %>% operator lets you chain together functions
It sends the result of one function to another function
Read %>% as “then”
```{r}
long_merged_energy %>% 
  filter(source == "geothermal") %>% 
  select(-datetime) %>% 
  mutate(log_usage = log(usage)) %>% 
  summarize(mean_log_usage = mean(log_usage, na.rm = T))
```

While piping, the piped dataframe is not changed!
To refer to the manipulated dataframe, use .

*Exercise*
**MAC --> CMD+SHIFT+M**
**PC --> CNTL+SHIFT+M**

Using the (wide) merged CA energy data merged_energy do the following with pipes:
1. Select variables that contain the word “hydro”
2. Create a new variable called total_hydro that is the sum of the retained hydro variables
3. Find the mean usage for total_hydro
```{r}
head(merged_energy)
```

```{r}
merged_energy %>% 
    select(contains("hydro")) %>% 
    mutate(total_hydro = rowSums(.,na.rm = T)) %>% 
    summarize(mean_hydro = mean(total_hydro, na.rm = T))
```


### group_by and summarize
perform operations by groups of observations
```{r}
long_merged_energy %>% 
  group_by(source) %>% 
  summarize(sum_usage = sum(usage, na.rm = T))
```

group_by versus for loops    
```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"))

gapminder %>% 
  group_by(year) %>% 
  summarize(mean_le = mean(lifeExp, na.rm = T),
            sd_lf = sd(lifeExp, na.rm = T))
```

Find the mean by source
Use your knowledge of dplyr to find the mean usage for small hydro, large hydro, biogas, and biomass
Start with either the wide or long merged dataset
```{r}
long_merged_energy %>% 
    group_by(source) %>% 
    # same as source == "small_hydro" | source == "large_hydro" ... 
    summarize(mean_usage = mean(usage, na.ram = T)) %>% 
    filter(source %in% c("small_hydro", "large_hydro", "biogas", "biomass"))
```

Alternative
```{r}
merged_energy %>% 
  select(datetime, contains("hydro"), contains("bio")) %>% 
  melt(id.vars = "datetime",
       variable.name = "source",
       value.name = "usage") %>% 
  group_by(source) %>% 
  summarize(mean_usage = mean(usage, na.rm = T))
```

Merging in dplyr with *_join functions
dplyr’s philosophy is to have more functions with fewer options

1. left_join - keeps all observations in the first dataframe
merge(all.x = T)
2. right_join - keeps all observations in the second dataframe
merge(all.y = T)
3. full_join - keeps all observations in both dataframes
merge(all = T)
4. inner_join - keeps only those observations that are matched in both datasets
merge(all = F)

Check dplyr cheatsheets!

### data table
```{r}
# import data.table library
library(data.table)

data_file <- here::here("data", "ca_energy_generation.csv")

# read in two versions of data, one as a data.frame and one as a data.table
generation_df <- read.csv(data_file, stringsAsFactors = F)

generation_dt <- fread(data_file)
```

```{r}
class(generation_df)
class(generation_dt)
```
```{r}
View(generation_df)
View(generation_dt)
generation_df
generation_dt
str(generation_df)
str(generation_dt)
```


**data table syntax**
think in terms of basic units -- rows, columns, and groups
data.table syntax provides placeholder for each of them

general form
DT[i,j,by] -- on which rows? what to do? grouped by what?

1. i: “on which rows” (row filtering)
```{r}
generation_dt[wind > 4400]
```

```{r}
generation_dt[wind > 4400 & mday(datetime) == 7]
```

*Exercise*

try it! select rows for which natural gas generation is less than or equal to 5,000 MW and large hydro generation is greater than 2,000 MW
```{r}
generation_dt[natural_gas <= 5000 & large_hydro > 2000]

```
bonus: select rows for which coal generation is greater than 10 MW and solar generation is greater than the median value of solar generation
```{r}
generation_dt[coal > 10 & solar > median(solar)]
```

2. j: “what to do” (column operations)
```{r}
generation_dt[,wind + solar]
```

```{r}
generation_dt[,3*wind + solar*biogas/2]
```

- New columns may be assigned in j or a new table can be created
```{r}
# := in place, add new columns in original table
generation_dt[,newcol := 3*wind + solar*biogas/2]
```

```{r}
# .() export data, create new table
generation_dt[,.(newcol = 3*wind + solar*biogas/2)]
```

- Columns may be deleted in j
```{r}
generation_dt[,newcol := NULL]
```

*Exercise*

try it! add a column called “total_hydro” that is the sum of the small_hydro and large_hydro columns
```{r}
generation_dt[,total_hydro := small_hydro + large_hydro]
```
try it! find the mean of the nuclear and biogas columns
```{r}
generation_dt[,.(mean(nuclear), mean(biogas))]
```
bonus: create a new table: for the hours when solar generation is zero, get the datetime and total_thermal (sum of natural gas and coal generation)
```{r}
generation_dt[solar == 0 ,.(datetime, total_thermal = natural_gas + coal)]
```

3. by: “grouped by what” (add grouping to any operation)

```{r}
generation_dt[,.(mean_nuc = mean(nuclear), mean_wind = mean(wind)), 
              by = mday(datetime)]
```


*Exercise*

try it! find the median solar generation by hour.
```{r}
generation_dt[,.(median_solar = median(solar)), by = hour(datetime)]
```
try it! for hours when the solar generation is greater than zero, find the maximum natural gas generation by day
```{r}
generation_dt[solar > 0,.(max_ng = max(natural_gas)), by = mday(datetime)]
```


### dplyr -> data.table

try it! Convert this dplyr syntax into data.table syntax (remember that this created the columns day, log_output, and per_output)
```{r}
long_ca_energy <- long_ca_energy %>%
  mutate(day = as_date(datetime),
         log_output = log(output)) %>%
  group_by(day) %>%
  mutate(total_daily_output = sum(output, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(per_output = output/total_daily_output)
```

becomes
```{r}
all_generation_long[,day := as_date(datetime)]
all_generation_long[,log_output := log(value)]
all_generation_long[,per_output := value/sum(value), by = day]
```

Or in one command
```{r}
all_generation_long[,`:=`(day2 = as_date(datetime), 
                          log_output2 = log(value), 
                          per_output2 = value/sum(value)), 
                    by = day]
```

### fread and fwrite
Similar to read.table but faster and more convenient










