---
title: "Day 3"
output: html_document
---

### ggplot
  
```{r}
source(here::here("data/day3_objects.R"))
```

```{r}
library(dplyr)
library(ggplot2)
```

*ggplot(data = <DATA FRAME>) + <GEOM_FUNCTION>(mapping = aes(<VARIABLES>))*
  
- data: a data frame, provided to the ggplot() function
- geometric objects: the objects/shapes that you want to plot, indicated through one of the many available geom functions, such as geom_point() or geom_hist()
- aesthetic mapping: the mapping from the data to the geometric objects, provided in an aes() function nested within the chosen geom function
- connected with the + operator

```{r}
ggplot(data = gapminder07) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp))
```

```{r}
ggplot(gapminder07) + 
  geom_point(aes(x = gdpPercap, y = lifeExp)) + 
  labs(title = "Relationship between life expectancy and GDP per capita in 2007", 
       x = "GDP per capita", y = "Life expectancy")
```

**Exercise**
Using the gapminder07 data, create a scatterplot of the natural log of gdpPercap as a function of the natural log of pop. Give it a title and axis labels.

Remember, you will need three functions: ggplot(), geom_point(), and labs().
```{r}
ggplot(gapminder07) + 
  geom_point(aes(x = log(pop), y = log(gdpPercap))) + 
  labs(title = "Relationship between GDP per capita and population in 2007", x = "logged GDP per capita", y = "logged life expectancy")
```

### dplyr and ggplot2

Scatterplots, generated with geom_point()
Column charts, generated with geom_col()

Plot a column chart of total energy generated over time.

1. Choose long-format generation data, i.e. long_gen.
2. Manipulate data frame to calculate total output per date-time.
3. Pipe manipulated data frame into plot.
4. Select geom_col() and map appropriate x and y variables.
```{r}
long_gen %>% 
    group_by(datetime) %>% 
    summarize(output = sum(output)) %>% 
    ggplot()+
    geom_col(aes(x = datetime, y = output)) +
    labs(title="Total energy generated, by hour", x = "Hour", y = "Output (MW)")
```


Plot a column chart hydroelectric power generated over time.
Hint: There are two types of hydroelectric sources in the data: large_hydro and small_hydro.
```{r}
long_gen %>% 
    filter(source %in% c("large_hydro", "small_hydro")) %>% 
    # filter(source=="large_hydro"|source=="small_hydro") %>%
    group_by(datetime) %>% 
    summarize(output = sum(output)) %>% 
    ggplot()+
    geom_col(aes(x = datetime, y = output))+
    labs(title="Total hydro power generated, by hour", x="Hour", y="Output (MW)")
```

Alternative
```{r}
generation %>% 
  mutate(hydro=large_hydro+small_hydro) %>%
  ggplot() + 
  geom_col(aes(x=datetime, y=hydro)) + 
  labs(title="Total hydro power generated, by hour", x="Hour", y="Output (MW)")
```


geom_line()
```{r}
imports %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=imports)) + 
  labs(title="Energy imports over time", x="Hour", y="Amount imported (MW)")
```

changing geom characteristics by modulate size= and col=.
```{r}
imports %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=imports), size=1.2, col="red") + 
  labs(title="Energy imports over time", x="Hour", y="Amount imported (MW)")
```

Area plots: geom_area()
```{r}
generation %>% 
  ggplot() + 
  geom_area(aes(x=datetime, y=wind), fill="darkblue") + 
  labs(title="Hourly wind power generation, Sept 3-9", x="Hour", y="Output (MW)")
```

Box plots: geom_boxplot()
25%, median, 75% --> interquartile range
```{r}
long_gen %>% 
  ggplot() + 
  geom_boxplot(aes(x=source, y=output)) + 
  labs(title="Amount of energy generated by each source, Sept 3-9", x="Source type", y = "Output (MW)")
```

Multiple geoms in one plot

Plot a line of large hydro generation over time, and a smoothed line of the same relationship on top of it.

Steps:

1. Supply the generation data to ggplot().
2. Create a line with geom_line() with appropriate x and y aesthetics. Make it turquoise for fun.
3. Create a smoothed line with geom_smooth() with the same x and y aesthetics.
4. geom_smooth() plots smoothed conditional means (estimated using a loess regression)
```{r}
generation %>% 
  ggplot() + 
  geom_line(aes(x=datetime, y=large_hydro), col="turquoise3") + 
  geom_smooth(aes(x=datetime, y=large_hydro)) + 
  labs(title="Hydroelectric (large) generation per hour, Sept 3-9", x="Hour", y="Output (MW)")
## `geom_smooth()` using method = 'loess' and formula 'y ~ x'
```

**Exercise 3: Total output per source**

Create a column chart that shows the total output per source.

1. Change the color of the columns to "darkred".
2. Add a horizontal line indicating the mean output across all sources. Use the cheatsheet to identify the geom function that you need.
3. Add a meaningful title and axis labels using labs().
```{r}
long_merged_energy %>% 
    group_by(source) %>% 
    summarize(output = sum(output)) %>% 
    ggplot() +
    geom_col(aes(x = source, y = output), fill = "darkred")+
    geom_hline(aes(yintercept = mean(output))) +
    geom_hline(aes(yintercept = median(output)))
    labs(title="Total output per energy source over Sept 3-9", y = "Output (MW)", x = "Source")
```


### scale
scales “control the details of how data values are translated to visual properties”

scale_alpha: control transparency settings
scale_color and scale_fill: control color palettes and other aspects of data mapped to colors
scale_x and scale_y: control the position of axis markers

Recreate line plot of imports over time, but label x-axis with hours rather than dates.

1. Supply imports data to ggplot().
2. Add a geom_line() layer and map appropriate x and y variables.
3. Add a scale_x_datetime() layer and use date_labels= and date_breaks= arguments to modify x-axis labels and breaks.
4. Add a labs() layer.
```{r}
imports %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=imports), col="red") + 
  scale_x_datetime(date_labels="%H:%M", date_breaks="12 hours") +
  labs(title="Energy imports over time in California", subtitle="Hourly data from September 3-9, 2018", x="Hour", y="Amount imported (MW)")
```


### Themes: presets
ggplot comes with several preset themes that can be added as layers, including:

theme_grey(): default
theme_bw(): strip colors, including grey gradients
theme_dark() and theme_light(): change the background of the coordinate system
theme_minimal() and theme_void(): see reference guide for details
```{r}
imports %>%
  ggplot() + geom_line(aes(x=datetime, y=imports), col="red") + 
  scale_x_datetime(date_labels="%H:%M", date_breaks="12 hours") +
  labs(title="Energy imports over time in California", subtitle="Hourly data from September 3-9, 2018", x="Hour", y="Amount imported (MW)") + 
  theme_minimal()
```

The theme() layer lets you modulate many components of the theme, including:

Angle, position, and other characteristics of axis labels (e.g. axis.labels.x =).
Angle, position, and other characteristics of legends (e.g. legend.position =).
Characteristics of the background of the plot (e.g. plot.background =).
For a full list of the components that can be modified, see the help file for ggplot2::theme().
```{r}
imports %>%
  ggplot() + geom_line(aes(x=datetime, y=imports), col="red") + 
  scale_x_datetime(date_labels="%H:%M", date_breaks="12 hours") +
  labs(title="Energy imports over time in California", subtitle="Hourly data from September 3-9, 2018", x="Hour", y="Amount imported (MW)") + 
  theme(axis.text.x=element_text(angle=45, hjust=1, size=12))
```

### Coordinate system adjustment
coord_flip() is the default Cartesian coordinate system used; by explicitly calling this function, you can change the limits of the x and y axes from their defaults
coord_fixed() sets a fixed aspect ratio between the x and y axes
coord_transform() lets you transform the Cartesian coordinates using functions like sqrt or log
coord_polar() changes the coordinate system to polar coordinates rather than a Cartesian system

**coord_flip() **
```{r}
long_gen %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date) %>% summarise(output=sum(output)) %>% 
  ggplot() + geom_col(aes(x=date, y=output)) + 
  labs(title="Total energy generated, by day", x="Day", y="Output (MW)")
```

after flipping
```{r}
long_gen %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date) %>% summarise(output=sum(output)) %>% 
  ggplot() + geom_col(aes(x=date, y=output)) + 
  labs(title="Total energy generated, by day", x="Day", y="Output (MW)") + 
  coord_flip()
```

### Visualizing grouped data
We often want to visualize multiple variables, or a variable that can be divided into groups, in one graphic.
There are two broad principles for visualizing grouped variables in R:
1. Identify a grouping variable in a geom() using the group = argument, and demarcate the distinct objects created per group in some way.
2. Use facets to generate geometric objects in separate coordinate systems for each group.

**Colors and fill**

Create a line plot of energy output over time, with separate lines for each source.

1. Supply the data frame long_merged_energy to ggplot().
2. Add a geom_line() layer and specify x=datetime and y=output.
3. Also specify that group=source and col=source.
4. Add a labs() layer.
```{r}
long_merged_energy %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=output, group=source, col=source)) + 
  labs(title="Output by energy source over time", subtitle="Hourly data from September 3-9, 2018", x="Hour", y="Output (MW)")
```

*Exercise 4: Colors and fill*

Create a line plot that compares generation of wind, solar, and geothermal energy over time.

Bonus: Set the line size to 1.5.
```{r}
long_merged_energy %>% 
    filter(source %in% c("wind", "solar", "geothermal")) %>% 
    # filter(source=="wind"|source=="solar"|source=="geothermal") %>%
    ggplot()+
    geom_line(aes(x = datetime, y=output, group = source, col = source), size = 1.5) +
    labs(title="Wind vs. Solar vs. Geothermal generation", subtitle="Hourly data from September 3-9, 2018", x="Hour", y="Output (MW)")
```

Customizing with scale_color layers
```{r}
long_merged_energy %>% filter(source=="wind"|source=="solar"|source=="geothermal") %>% 
  ggplot() + 
  geom_line(aes(x=datetime, y=output, group=source, col=source), size=1.5) + 
  scale_color_brewer(palette="Accent", name="Energy source") + 
  labs(title="Wind vs. Solar vs. Geothermal generation", subtitle="Hourly data from September 3-9, 2018", x="Hour", y="Output (MW)")
```

Colors and fill

col= is used to color objects like lines and points
fill= is used to color objects like columns and histograms
col= will create colored outlines for such objects
After some trial and error, this will become intuitive


### Position adjustment
Example: Energy use by day

Create a column chart of energy use by day, grouped by source.

1. Modify long_merged_energy to summarize output by date and source.
    (1) Use mutate() to create a date variable from datetime.
    (2)Use group_by() and summarize() to calculate total output per date per source.
2. Supply modified data frame to ggplot()
3. Add a geom_col() layer, supply x and y aesthetics.
4. Specify group=source and fill=source.
5. Add a labs() layer.

```{r}
long_merged_energy %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date, source) %>% 
  summarize(output=sum(output)) %>% 
  ggplot() + 
  geom_col(aes(x=date, y=output, group=source, fill=source)) + 
  labs(title="Energy use by day", x="Day", y="Output (MW)")
```

?? But what if our main goal is to compare trends across sources?
We can use position="dodge" in geom_col() ‘unstack’ the columns for each source.

*pos dodge*
```{r}
long_merged_energy %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date, source) %>% 
  summarize(output=sum(output)) %>% 
  ggplot() + 
  geom_col(aes(x=date, y=output, group=source, fill=source), position="dodge") + 
  labs(title="Energy use by day", x="Day", y="Output (MW)")
```


?? What if our goal is to compare what portion of output per day each source comprises?
We can use position="fill" to normalize height of stacked columns.

*pos fill*
```{r}
long_merged_energy %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date, source) %>% 
  summarize(output=sum(output)) %>% 
  ggplot() + 
  geom_col(aes(x=date, y=output, group=source, fill=source), position="fill") + 
  labs(title="Energy use by day", x="Day", y="Output (MW)")
```


### Shapes and linetypes
Colors/fill and position adjustment are the most common and intuitive way to distinguish geometric objects from each other.
Shapes and linetypes are two other ways.
The idea here is to modify what the object actually looks like.

Exercise 3: Total output per source
long_merged_energy %>%
  group_by(source) %>% 
  summarise(output=sum(output)) %>%
  ggplot() + 
  geom_col(aes(x=source, y=output), fill="darkred") + 
  geom_hline(aes(yintercept=mean(output))) + 
  labs(title="Total output per energy source over Sept 3-9", y="Output (MW)", x="Source")


Example: coord_flip()
long_gen %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date) %>% summarise(output=sum(output)) %>% 
  ggplot() + geom_col(aes(x=date, y=output)) + 
  labs(title="Total energy generated, by day", x="Day", y="Output (MW)") + 
  coord_flip()


Stats
There are several stat functions in ggplot that enable you to conduct statistical transformations of your data prior to plotting.
stat and geom layers can be used interchangeably, as each stat layer has a geom argument and vice versa.
You may want to use stat layers or arguments to override default settings, or to use transformed versions of a variable in your plot.
For more, see Chapter 3.7 of the R for data science online textbook.
Example: Energy use by day (pos fill)
long_merged_energy %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date, source) %>% 
  summarize(output=sum(output)) %>% 
  ggplot() + 
  geom_col(aes(x=date, y=output, group=source, fill=source), position="fill") + 
  labs(title="Energy use by day", x="Day", y="Output (MW)")


Shapes and linetypes
Colors/fill and position adjustment are the most common and intuitive way to distinguish geometric objects from each other.
Shapes and linetypes are two other ways.
The idea here is to modify what the object actually looks like.

*Example: Energy source group, by day*

Prepare data frame that summarizes output by date and “group” from regroup.
Supply modified data frame to ggplot().
Two choices!
Add geom_line() and geom_point() with shape=group.
Add geom_line() with linetype=group.
```{r}
# Prepare data
long_merged_energy_regroup <- long_merged_energy %>%
  rename(type = source) %>% 
  merge(regroup, by = "type") %>% 
  mutate(date=lubridate::date(datetime)) %>% 
  group_by(date, group) %>%
  summarise(output=sum(output))

# Take a look at our prepared data
head(long_merged_energy_regroup)
```


```{r}
long_merged_energy_regroup %>% 
  ggplot() + 
  geom_line(aes(x=date, y=output, group=group, col=group), size=0.8) +
  geom_point(aes(x=date, y=output, group=group, shape=group)) + 
  labs(title="Output by source group over time", subtitle="Data collected during September 3-9, 2018", x="Date", y="Output (MW)")
```


```{r}
long_merged_energy_regroup %>% 
  ggplot() + 
  geom_line(aes(x=date, y=output, group=group, linetype=group), size=1) +
  labs(title="Output by source group over time", subtitle="Data collected during September 3-9, 2018", x="Date", y="Output (MW)")
```


Sizes and alpha
size= and alpha= inside aes() modulate the size and transparency of geom objects based on some data value.
They are particularly useful when you want to distinguish objects based on some continuous variable.
Let’s return to the Gapminder data for an example.

```{r}
gapminder07 %>% 
  ggplot() + 
  geom_point(aes(x=log(gdpPercap), y=lifeExp, size=pop, col=continent)) + 
  scale_size_continuous(name="Population") + scale_color_discrete(name="Continent") +
  labs(title="Life expectancy as a function of GDP per capita in 2007", x="Logged GDP per capita", y="Life expectancy")
```


*Exercise 5: Average hourly output by source*

Exercise 5: Average hourly output by source

Visualize the average output for each hour of the day, grouped by source.

You need to identify the output per source per hour (e.g. 01:00, 02:00, etc) averaged over all days.
You will need to prepare your data using both dplyr and lubridate functions.
You can choose which geom(s) to use, and how to demarcate groups.
Bonus: use a scale layer to set a color palette (try "Set3") and change the legend name.
Remember to add labs()!

```{r}
ex5 <- long_merged_energy %>% 
  mutate(hour=lubridate::hour(datetime)) %>% 
  group_by(hour, source) %>% 
  summarize(output=sum(output)) %>%
  ggplot() + 
  geom_area(aes(x=hour, y=output, fill=factor(source))) + 
  scale_fill_brewer(palette="Set3", name="Source") + 
  labs(title="Average hourly output by source", 
       subtitle="Data collected during September 3-9",
       x="Hour of the day", y="Output (MW)") + 
  theme_bw()
ex5
```

*Example: Comparing generation patterns*

Compare energy generation over time, across sources.

How do we do this? Using what we’ve learned so far:

Supply long_gen to ggplot().
Add a geom_line() layer, setting x=datetime and y=output in aes().
Let’s set group=source and col=source in aes().
```{r}
long_gen %>%
  ggplot() + 
  geom_line(aes(x=datetime, y=output, group=source, col=source), size=1) + 
  labs(title="Generation over time, by energy source", subtitle="Hourly data from September 3-9, 2018", x="Hour", y="Output (MW)")
```

*Example: Comparing generation patterns*
Is this a helpful plot?

Not if our goal is to compare the patterns of each source.
It’s too noisy!
Instead of setting col= in aes(), let’s add a facet_wrap() layer.
Specifically, facet_wrap(~source), i.e. tell ggplot to “facet by source”.
```{r}
long_gen %>% 
  ggplot() + 
  geom_line(aes(x = datetime, y = output)) + 
  facet_wrap(~source) + 
  labs(title="Generation over time, by energy source", subtitle="Hourly data from September 3-9, 2018", x="Hour", y="Output (MW)")
```


```{r}
long_gen %>% 
  ggplot() + 
  geom_line(aes(x = datetime, y = output)) + 
  facet_wrap(~source, scales="free") + 
  labs(title="Generation over time, by energy source", subtitle="Hourly data from Sep 3-9, 2018", x = "Hour", y = "Output (MW)")
```

*Exercise 6: Facets*

Compare generation patterns for each source in facets. Color the lines using the “group” variable in regroup.

You will need to prepare your data. Think about the variables that you need: source/type, output, and group.
When you pipe modified data into ggplot, remember that you are grouping in two ways: source/type through facets, and group through color.
```{r}
long_gen_regroup <- long_gen %>%
  rename(type = source) %>% 
  merge(regroup, by="type")

head(long_gen_regroup)
```

```{r}
long_gen_regroup %>% ggplot() + 
  geom_line(aes(x=datetime, y=output, group=group, col=group), size=1) + 
  scale_color_brewer(palette="Set1", name="Type of energy source") +
  facet_wrap(~type, scales="free") + 
  labs(title="Generation over time, by energy source", subtitle="Hourly data from September 3-9, 2018", x="Hour", y="Output (MW)") + 
  theme(legend.position = "bottom")
```

### Saving images
Three ways to save images created by ggplot:

Assign the result of your ggplot code to a named object, which can be saved like other objects in your R workspace
Use the “save” icon in the Plots tab of RStudio
Add the function ggsave() as a layer to your plotting code










