---
title: "Rmd Exercise Template"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The question you are answering

You use the `##` to create a header. Use the header to indicate the question you are answering. After the header, in this space here, annotate your answer, telling us what you did.

```{r chunk name}
# Here is where your R code goes. ``` opens the code chunk, followed by {r}. Then ``` closes the chunk. When you 'knit' the file, the codes will evaluate.

#dat <- read.csv("path/to/data/data.csv")
```

## Including Plots

You can also embed plots, for example:

```{r another chunk name}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


## Load the Data
Load the `gapminder` dataset.
```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```


## If Statement
Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the any function.
```{r}
year <- 2002
if(any(gapminder$year == year)){
    print(paste("Records from",year,"were found"))
}else{
    print("Nothing found")
}
```
```{r}
year <- 2012
if(any(gapminder$year == year)){
    print(paste("Records from",year,"were found"))
}else{
    print("Nothing found")
}
```

##Loop and If Statements
Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset
```{r}
countries <- unique(gapminder$country)
for(i in countries){
    if(mean(gapminder$pop[gapminder$country == i]) < mean(gapminder$pop)){
        mean_le <- mean(gapminder$lifeExp[gapminder$country == i])
        print(paste("Mean life expectancy for", i, "is", mean_le))
    }
}
```



Write a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.
```{r}
for (i in unique(gapminder$continent)){
    mean_le <- mean(gapminder$lifeExp[gapminder$continent == i])
    if(mean_le < 50){
        print(paste("Mean Life Expectancy in", i, "is smaller than 50"))
    }
    if(mean_le > 50 & mean_le < 70){
        print(paste("Mean Life Expectancy in", i, "is between 50 and 70"))
    }
    if(mean_le > 70){
        print(paste("Mean Life Expectancy in", i, "is greater than 70"))
    }
}
```


### Writing Functions
Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset. Hint: Use mode() or class() to get the class of the data in each column. Remember that names() or colnames() returns the name of the columns in a dataset.
```{r}
data_frame_name_class <- function(df){
    col <- names(df)
    for(i in col){
        print(paste(i, ":", mode(df[,i])))
    }
}
data_frame_name_class(gapminder)
```


















