---
title: "Rmd Exercise Template"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The question you are answering

You use the `##` to create a header. Use the header to indicate the question you are answering. After the header, in this space here, annotate your answer, telling us what you did.

```{r chunk name}
# Here is where your R code goes. ``` opens the code chunk, followed by {r}. Then ``` closes the chunk. When you 'knit' the file, the codes will evaluate.

#dat <- read.csv("path/to/data/data.csv")
```

## Including Plots

You can also embed plots, for example:

```{r another chunk name}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


## Load the Data
Load the `gapminder` dataset.
```{r}
gapminder <- read.csv(here::here("data/gapminder5.csv"), stringsAsFactors=FALSE)
```


## If Statement
Use an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. Now do the same for 2012.

Hint: use the any function.
```{r}
year <- 2002
if(any(gapminder$year == year)){
    print(paste("Records from",year,"were found"))
}else{
    print("Nothing found")
}
```
```{r}
year <- 2012
if(any(gapminder$year == year)){
    print(paste("Records from",year,"were found"))
}else{
    print("Nothing found")
}
```

##Loop and If Statements
Write a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset
```{r}
countries <- unique(gapminder$country)
for(i in countries){
    if(mean(gapminder$pop[gapminder$country == i]) < mean(gapminder$pop)){
        mean_le <- mean(gapminder$lifeExp[gapminder$country == i])
        print(paste("Mean life expectancy for", i, "is", mean_le))
    }
}
```



Write a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.
```{r}
for (i in unique(gapminder$continent)){
    mean_le <- mean(gapminder$lifeExp[gapminder$continent == i])
    if(mean_le < 50){
        print(paste("Mean Life Expectancy in", i, "is smaller than 50"))
    }
    if(mean_le > 50 & mean_le < 70){
        print(paste("Mean Life Expectancy in", i, "is between 50 and 70"))
    }
    if(mean_le > 70){
        print(paste("Mean Life Expectancy in", i, "is greater than 70"))
    }
}
```


### Writing Functions
Create a function that given a data frame will print the name of each column and the class of data it contains. Use the gapminder dataset. Hint: Use mode() or class() to get the class of the data in each column. Remember that names() or colnames() returns the name of the columns in a dataset.
```{r}
data_frame_name_class <- function(df){
    col <- names(df)
    for(i in col){
        print(paste(i, ":", mode(df[,i]))) # class[,i]
    }
}
data_frame_name_class(gapminder)
```


Create a function that given a vector will print the mean and the standard deviation of a vector, it will optionally also print the median. Hint: include an argument that takes a boolean (TRUE/FALSE) operator and then include an if statement.
```{r}
vec_mean_sd <- function(vec, flag = FALSE){
    print(paste("Mean:", mean(vec)))
    print(paste("Standard Deviation:", sd(vec)))
    if (flag == TRUE){
        print(paste("Median:", median(vec)))
    }
}
vec_mean_sd(gapminder$lifeExp, flag = TRUE)
```
```{r}
vec_mean_sd(gapminder$lifeExp, flag = FALSE)
```

### Analyzing the relationship between GDP per capita and life expectancy
Use what you've learned so far to answer the following questions using the gapminder dataset. Be sure to include some visualizations!

1. What is the relationship between GDP per capita and life expectancy? Does this relationship change over time? (Hint: Use the natural log of both variables.)
```{r}
viz_lm <-
  function(df, dv, iv, year) {
    dat <- df[df[["year"]] == year, ]
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x)
    plot(y ~ x, main = year,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'blue')
  }
```

```{r}
for (i in years) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)
}

```

2. Does the relationship between GDP per capita and life expectacy vary by continent? Make sure you divide the Americas into North and South America.

```{r}
viz_lm <-
  function(df, dv, iv, continent) {
    dat <- df[df[["continent"]] == continent, ]
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x)
    plot(y ~ x, main = continent,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'blue')
  }
```

```{r}
for (i in continent) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)
}
```

```












